Distinguishing Spin-Aligned and Isotropic Black Hole Populations With Gravitational Waves

### 2017-06-04 paper/AlignedIsotropicGW.tex
Some typos fixed, some references eliminated.

\begin{abstract}
  The first direct detections of gravitational
  waves\cite{2016PhRvL.116f1102A,2016PhRvL.116x1103A,O1-BBH,PhysRevLett.118.221101}
  from merging binary black holes open a unique window into the binary
  black hole formation environment.  One promising environmental
  signature is the angular distribution of the black hole spins;
  systems formed through dynamical interactions among already-compact
  objects are expected to have isotropic spin
  orientations\cite{SigurdssonHernquist:1993,1993Natur.364..421K,PZMcMillan:2000,Rodriguez:2015,Stone:2016,2016ApJ...832L...2R}
  whereas binaries formed from pairs of stars born together are more
  likely to have spins preferentially aligned with the binary orbital
  angular
  momentum\cite{TutukovYungelson:1993,Lipunov:1997,2016Natur.534..512B,Stevenson:2017,MandeldeMink:2016,Marchant:2016}.
  We consider existing gravitational wave measurements of the binary
  effective spin, the best-measured combination of spin
  parameters\cite{2016PhRvL.116x1102A,O1-BBH,PhysRevLett.118.221101},
  in the four likely binary black hole detections GW150914, LVT151012,
  GW151226, and GW170104.  If binary black hole spin magnitudes extend
  to high values we show that the data exhibit a
  $\OOneSigmaIsoAligned\sigma$ ($\OOneOddsIsoAligned$ odds
  ratio\footnote{An odds ratio of $r$ with $r \ll 1$ is equivalent to
    $x \sigma$ with $x = \Phi^{-1}\left( 1 - r/2 \right)$, where
    $\Phi$ is the unit normal CDF.})  preference for an isotropic
  angular distribution over an aligned one.  By considering the effect
  of 10 additional detections\cite{2016ApJ...833L...1A}, we show that
  such an augmented data set would enable in most cases a prefernce
  stronger than $5\sigma$ ($2.9 \times 10^{-7}$ odds ratio).  The
  existing preference for either an isotropic spin distribution or low
  spin magnitudes for the observed systems will be confirmed (or
  overturned) confidently in the near future.
\end{abstract}

@@

Following the detection of a merging binary black hole system,
parameter estimation tools\cite{2015PhRvD..91d2003V} compare model
gravitational waveforms against the observed data to obtain a
posterior distribution on the parameters that describe the compact
binary source.  The spin parameter with the largest effect on
waveforms, and a correspondingly tight constraint from the
data\cite{2016PhRvL.116x1102A}, is a mass-weighted combination of the
components of the dimensionless spin vectors of the two black holes
that are aligned with the orbital axis, the ``effective spin,''
$-1 < \chieff < 1$ (see Methods Section
\ref{methsec:chieff-spin-magnitude}).

@@

We fit hierarchical models of the four existing LIGO O1 and GW170104
observations using these six different, zero-parameter population
distributions (see Methods Section \ref{sec:hierarchical}).  We also
fit three mixture models for the population, where the spin magnitude
distribution is fixed but the angular distribution is a weighted sum
of the isotropic and aligned distribution.  The evidence, or marginal
likelihood, for each of the models is shown in Figure
\ref{fig:O1-odds}.  For all three magnitude distributions, the mixture
models' posterior on the mixing fraction peaks at 100\% isotropic,
which explains why the zero-parameter, pure-isotropic models are
preferred over the single-parameter mixture models for every magnitude
distribution with this data set.  Not surprisingly, given the small
$\chieff$ values in the three detected systems, the most-favoured
model among those with an isotropic angular distribution has the
``low'' magnitude distribution; the most favoured model among those
with an aligned distribution also has the ``low'' magnitude
distribution.  The odds ratio between the ``low'' aligned and ``low''
isotropic models is $\OOneOddsIsoAligned$, or
$\OOneSigmaIsoAligned\sigma$; thus the data favour isotropic spins
among our suite of models.  While the data favour spin amplitude
distributions with small spin magnitudes, note that a model with all
binary black hole systems having zero spin is ruled out by the
GW151226 measurements, which bound at least one black hole to have
spin magnitude $\geq 0.2$ at 99\%
credibility\cite{2016PhRvL.116x1103A}.

@@

Observational data on spin magnitudes in black hole systems is
sparse\cite{2015PhR...548....1M}.  Most of the systems studied are
low-mass X-ray binaries rather than the high-mass X-ray binaries that
are likely to be the progenitors of double black hole binaries.  In
addition, there are substantial systematic errors that can complicate
these analyses\cite{2015PhR...548....1M} and selection effects could
yield a biased distribution. Nonetheless, if we take the reported spin
magnitudes as representative then we find that there is a preference
for high spins; for example, 14 of the 19 systems with reported spins
have dimensionless spin parameters in excess of 0.5.  It is usually
argued that the masses and spin parameters of stellar-mass black holes
are unlikely to be altered significantly by
accretion\cite{1999MNRAS.305..654K}, but this may not be true for all
systems\cite{2003MNRAS.341..385P,2015ApJ...800...17F}.  Thus the
current spin parameters are probably close to their values upon core
collapse, at least in high-mass X-ray binaries.  However, the specific
processes involved in the production of black hole binaries from
isolated binaries could alter the spin magnitude distribution of those
holes relative to the X-ray binary systems; for example, close tidal
interactions could spin up the core, or stripping of the envelope
could reduce the available angular
momentum\cite{2016MNRAS.462..844K,2017arXiv170200885Z,2017arXiv170203952H}.

>>>>

\begin{abstract}
  The first direct detections of gravitational
  waves\cite{2016PhRvL.116f1102A,2016PhRvL.116x1103A,O1-BBH,PhysRevLett.118.221101}
  from merging binary black holes open a unique window into the binary
  black hole formation environment.  One promising environmental
  signature is the angular distribution of the black hole spins;
  systems formed through dynamical interactions among already-compact
  objects are expected to have isotropic spin
  orientations\cite{SigurdssonHernquist:1993,1993Natur.364..421K,PZMcMillan:2000,Rodriguez:2015,Stone:2016,2016ApJ...832L...2R}
  whereas binaries formed from pairs of stars born together are more
  likely to have spins preferentially aligned with the binary orbital
  angular
  momentum\cite{TutukovYungelson:1993,Lipunov:1997,2016Natur.534..512B,Stevenson:2017,MandeldeMink:2016,Marchant:2016}.
  We consider existing gravitational wave measurements of the binary
  effective spin, the best-measured combination of spin
  parameters\cite{O1-BBH,PhysRevLett.118.221101},
  in the four likely binary black hole detections GW150914, LVT151012,
  GW151226, and GW170104.  If binary black hole spin magnitudes extend
  to high values we show that the data exhibit a
  $\OOneSigmaIsoAligned\sigma$ ($\OOneOddsIsoAligned$ odds
  ratio\footnote{An odds ratio of $r$ with $r \ll 1$ is equivalent to
    $x \sigma$ with $x = \Phi^{-1}\left( 1 - r/2 \right)$, where
    $\Phi$ is the unit normal CDF.})  preference for an isotropic
  angular distribution over an aligned one.  By considering the effect
  of 10 additional detections\cite{2016ApJ...833L...1A}, we show that
  such an augmented data set would enable in most cases a prefernce
  stronger than $5\sigma$ ($2.9 \times 10^{-7}$ odds ratio).  The
  existing preference for either an isotropic spin distribution or low
  spin magnitudes for the observed systems will be confirmed (or
  overturned) confidently in the near future.
\end{abstract}

@@

Following the detection of a merging binary black hole system,
parameter estimation tools\cite{2015PhRvD..91d2003V} compare model
gravitational waveforms against the observed data to obtain a
posterior distribution on the parameters that describe the compact
binary source.  The spin parameter with the largest effect on
waveforms, and a correspondingly tight constraint from the
data\cite{O1-BBH}, is a mass-weighted combination of the components of
the dimensionless spin vectors of the two black holes that are aligned
with the orbital axis, the ``effective spin,'' $-1 < \chieff < 1$ (see
Methods Section \ref{methsec:chieff-spin-magnitude}).

@@

We fit hierarchical models of the three existing LIGO O1 and GW170104
observations using these six different, zero-parameter population
distributions (see Methods Section \ref{sec:hierarchical}).  We also
fit three mixture models for the population, where the spin magnitude
distribution is fixed but the angular distribution is a weighted sum
of the isotropic and aligned distribution.  The evidence, or marginal
likelihood, for each of the models is shown in Figure
\ref{fig:O1-odds}.  For all three magnitude distributions, the mixture
models' posterior on the mixing fraction peaks at 100\% isotropic,
which explains why the zero-parameter, pure-isotropic models are
preferred over the single-parameter mixture models for every magnitude
distribution with this data set.  Not surprisingly, given the small
$\chieff$ values in the three detected systems, the most-favoured
model among those with an isotropic angular distribution has the
``low'' magnitude distribution; the most favoured model among those
with an aligned distribution also has the ``low'' magnitude
distribution.  The odds ratio between the ``low'' aligned and ``low''
isotropic models is $\OOneOddsIsoAligned$, or
$\OOneSigmaIsoAligned\sigma$; thus the data favour isotropic spins
among our suite of models.  While the data favour spin amplitude
distributions with small spin magnitudes, note that a model with all
binary black hole systems having zero spin is ruled out by the
GW151226 measurements, which bound at least one black hole to have
spin magnitude $\geq 0.2$ at 99\%
credibility\cite{2016PhRvL.116x1103A}.

@@

Observational data on spin magnitudes in black hole systems is
sparse\cite{2015PhR...548....1M}.  Most of the systems studied are
low-mass X-ray binaries rather than the high-mass X-ray binaries that
are likely to be the progenitors of double black hole binaries.  In
addition, there are substantial systematic errors that can complicate
these analyses\cite{2015PhR...548....1M} and selection effects could
yield a biased distribution. Nonetheless, if we take the reported spin
magnitudes as representative then we find that there is a preference
for high spins; for example, 14 of the 19 systems with reported spins
have dimensionless spin parameters in excess of 0.5.  It is usually
argued that the masses and spin parameters of stellar-mass black holes
are unlikely to be altered significantly by
accretion\cite{1999MNRAS.305..654K}, but this may not be true for all
systems\cite{2003MNRAS.341..385P,2015ApJ...800...17F}.  Thus the
current spin parameters are probably close to their values upon core
collapse, at least in high-mass X-ray binaries.  However, the specific
processes involved in the production of black hole binaries from
isolated binaries could alter the spin magnitude distribution of those
holes relative to the X-ray binary systems; for example, close tidal
interactions could spin up the core, or stripping of the envelope
could reduce the available angular
momentum\cite{2016MNRAS.462..844K,2017arXiv170203952H}.
